---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
# Histograms for variables in 2020
variables <- c("ndvi_20_med", "maxtemp_20_med", "mintemp_20_med", "rain_20_sum", "pop_20", "water_20_pct", "elev_med")

par(mfrow = c(3, 3))  # Layout for multiple plots
for (var in variables) {
  hist(lab6_data[[var]], main = paste("Histogram of", var), xlab = var, col = "lightblue", border = "white")
}

```

{Your assessment of the variables' normality goes here.}
NDVI (`ndvi_20_med`) is heavily concentrated at lower values, which suggests a strong right-skew.
Maximum temperature (`maxtemp_20_med`) is strongly left_skewed with highly concentrated higher values.
On the other hand, minimum temperature (`mintemp_20_med`) display distributions closer to normal, appearing relatively symmetric around their peaks.
Rainfall (`rain_20_sum`) is highly right-skewed, with most areas experiencing low rainfall, while a few regions have extremely high values.
Population (`pop_20`) shows extreme skewness, with the majority of hexagons having very low population counts, and only a few having large populations.
Water percentage (`water_20_pct`) is also strongly right-skewed, with most areas having little to no water coverage.
Elevation (`elev_med`) is slightly right-skewed, with most values concentrated in the lower to mid-elevation range but less extreme than other variables like population or rainfall.

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
library(tmap)

# Set tmap mode to "plot" for static maps
tmap_mode("plot")

# Create individual maps using Jenks natural breaks
map_ndvi <- tm_shape(lab6_data) +
  tm_polygons("ndvi_20_med", style = "jenks", palette = "YlGn", title = "NDVI (2020)") +
  tm_layout(title = "NDVI", legend.outside = TRUE)

map_maxtemp <- tm_shape(lab6_data) +
  tm_polygons("maxtemp_20_med", style = "jenks", palette = "Reds", title = "Max Temp (2020)") +
  tm_layout(title = "Max Temperature", legend.outside = TRUE)

map_mintemp <- tm_shape(lab6_data) +
  tm_polygons("mintemp_20_med", style = "jenks", palette = "Blues", title = "Min Temp (2020)") +
  tm_layout(title = "Min Temperature", legend.outside = TRUE)

map_rain <- tm_shape(lab6_data) +
  tm_polygons("rain_20_sum", style = "jenks", palette = "PuBuGn", title = "Rainfall (2020)") +
  tm_layout(title = "Rainfall", legend.outside = TRUE)

map_population <- tm_shape(lab6_data) +
  tm_polygons("pop_20", style = "jenks", palette = "Purples", title = "Population (2020)") +
  tm_layout(title = "Population", legend.outside = TRUE)

map_water <- tm_shape(lab6_data) +
  tm_polygons("water_20_pct", style = "jenks", palette = "GnBu", title = "Water (%)") +
  tm_layout(title = "Water Percentage", legend.outside = TRUE)

map_elevation <- tm_shape(lab6_data) +
  tm_polygons("elev_med", style = "jenks", palette = "Oranges", title = "Elevation") +
  tm_layout(title = "Elevation", legend.outside = TRUE)

# Arrange all maps in a single figure
tmap_arrange(map_ndvi, map_maxtemp, map_mintemp, map_rain, map_population, map_water, map_elevation, ncol = 2)


```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

{Your response here.}
The spatial distribution of NDVI in 2020 reveals higher values in the northeastern and southeastern regions of Australia, which coincide with areas of higher rainfall, indicating a strong positive association. In contrast, central and western regions with higher maximum and minimum temperatures show significantly lower NDVI values, suggesting a negative relationship with max and min temperature. Population and elevation exhibit weaker associations, as NDVI is primarily driven by climatic factors like rainfall and temperature rather than human density or terrain.

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
# List of independent variables
independent_vars <- c("maxtemp_20_med", "mintemp_20_med", "rain_20_sum", "pop_20", "water_20_pct", "elev_med")

# Create univariate regression models and store them
univariate_models <- lapply(independent_vars, function(var) {
  formula <- as.formula(paste("ndvi_20_med ~", var))
  lm(formula, data = lab6_data)
})

# Print summaries of each model
model_summaries <- lapply(univariate_models, summary)

# Display summaries for interpretation
names(model_summaries) <- independent_vars
model_summaries

```

{Discussion here}
1. Maximum temperature: the results indicate a significant negative impact on the dependent variable. Specifically, for each one-unit increase in maximum temperature, the dependent variable decreases by approximately 0.0131 units on average. This relationship is highly significant (p<2e−16), suggesting that maximum temperature plays an important role in explaining variations in the dependent variable. The model explains 20.55% of the variance.

2. Minimum Temperature: the analysis also reveals a significant negative relationship. Each one-unit increase in minimum temperature results in an average decrease of 0.0123 units in the dependent variable. This relationship is statistically significant (p<2e−16), though the model explains a smaller portion of the variance, at 14.06%.

3. Rainfall: the results show a significant positive effect on the dependent variable. For every one-unit increase in rainfall, the dependent variable increases by approximately 0.0000009124 units on average. Although the effect size is small, the relationship is highly significant (p<2e−16), and this model demonstrates a strong explanatory power, accounting for 42.65% of the variance.

4. Population: population has a significant positive influence on the dependent variable. Each one-unit increase in population corresponds to an average increase of 0.0000015 units in the dependent variable. This relationship is highly significant (p<2e−16), though the model’s explanatory power is limited, explaining only 12.16% of the variance.

5. Water Percentage: the results suggest that the variable does not have a significant relationship with the dependent variable (p=0.249). While the coefficient indicates a potential negative trend (a decrease of 0.1783 units in the dependent variable for every 1% increase in water coverage), this effect is not statistically significant, and the model explains only 0.05% of the variance.

6. Elevation: indicates a significant positive effect on the dependent variable. For each one-unit increase in elevation, the dependent variable increases by approximately 0.000179 units on average. This relationship is statistically significant (p=1.14e−9), but the model explains only a small portion of the variance, at 4.93%.

In conclusion, rainfall and temperature are the most significant factors influencing NDVI, with rainfall playing the dominant role. Other variables, such as population and elevation, have weaker effects, while water percentage shows no meaningful relationship with vegetation.

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
# Multivariate regression model with selected variables
multi_model <- lm(ndvi_20_med ~ maxtemp_20_med + rain_20_sum + pop_20 + elev_med, data = lab6_data)

# Summary of the multivariate model
summary(multi_model)

```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

{Response goes here}
The multivariate model reveals that rainfall and maximum temperature are the most important predictors of NDVI, with rainfall showing a positive association (\(8.469 \times 10^{-7}\)) and maximum temperature a negative association (\(-0.0117\)). This indicates that areas with higher rainfall have more vegetation, while higher temperatures reduce vegetation levels. Population and elevation also contribute positively to NDVI, though their effects are smaller. Population increases NDVI slightly (\(2.862 \times 10^{-7}\) per unit), while elevation has a weak positive impact (\(0.0001233\) per meter). Compared to the univariate models, the direction and significance of all predictors remain consistent, but the magnitudes of their effects decrease slightly due to the inclusion of additional variables and shared variance. The model has a high adjusted \(R^2\) of 63.76%, explaining most of the variability in NDVI, and an F-statistic of 315.5 (p < 2.2e-16), confirming its strong overall fit.

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
# Load necessary libraries
library(tmap)
library(spdep)

# Calculate residuals from the model
lab6_data$residuals <- residuals(multi_model)

# Plot a histogram of residuals
par(mfrow = c(1, 1))  # Single plot layout
hist(
  lab6_data$residuals,
  main = "Histogram of Residuals",
  xlab = "Residuals",
  col = "lightblue",
  border = "white"
)

# Plot the spatial distribution of residuals
tm_shape(lab6_data) +
  tm_polygons(
    "residuals",
    style = "jenks",
    palette = "RdBu",
    title = "Residuals",
    midpoint = 0
  ) +
  tm_layout(
    main.title = "Spatial Distribution of Residuals",
    legend.outside = TRUE,
    legend.width = 2
  )

# Moran's I Test for spatial autocorrelation
# Extract coordinates
coords <- st_coordinates(st_centroid(lab6_data))  # Use centroids for spatial coordinates
nb <- knn2nb(knearneigh(coords, k = 5))  # Use 5 nearest neighbors
lw <- nb2listw(nb, style = "W")  # Convert neighbors to a weights list

# Perform Moran's I test
moran_test <- moran.test(lab6_data$residuals, lw)

# Print Moran's I results
print(moran_test)
```

{Summary goes here}
The histogram of residuals shows that the residuals are approximately bell-shaped, but there is a slight right skew. Most of the residuals are centered around 0, with very few extreme values on either side. This indicates that the residuals are reasonably close to normal, but not perfectly distributed.
And the results of Moran's I test show that the residuals exhibit significant positive spatial autocorrelation. The Moran I statistic of 0.24 suggests that similar residuals tend to cluster together in space. The extremely low p-value (< 2.2e-16) confirms that this spatial clustering is statistically significant.

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
library(car)
library(lmtest)

# Calculate VIF for the multivariate model
vif_values <- vif(multi_model)
print(vif_values)

# Check for heteroskedasticity with the Breusch-Pagan test
bp_test <- bptest(multi_model)
print(bp_test)

# Plot residuals vs. fitted values for a visual check of heteroskedasticity
plot(
  fitted(multi_model),
  residuals(multi_model),
  main = "Residuals vs. Fitted Values",
  xlab = "Fitted Values",
  ylab = "Residuals",
  col = "blue",
  pch = 16
)
abline(h = 0, col = "red")
```

{Interpretation goes here}
The Variance Inflation Factor (VIF) values for all variables in the model are below 1.2, indicating very low multicollinearity. This suggests that the independent variables are not highly correlated, and their individual effects can be reliably estimated. 

The results of the Breusch-Pagan test (BP = 122.21, df = 4, p-value < 2.2e-16) indicate significant heteroskedasticity in the model residuals, as the p-value is well below 0.05, leading us to reject the null hypothesis of homoskedasticity. This suggests that the variance of residuals is not constant across fitted values. The residuals vs. fitted values plot visually confirm the presence of heteroskedasticity. 

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._
The regression model shows that rainfall and maximum temperature are the most significant predictors of NDVI, with rainfall positively and temperature negatively associated with vegetation. The adjusted $R^2$ value of 63.76% indicates a strong model fit, and the F-statistic confirms its significance. In addition, the histogram of residuals suggests near-normality and the Moran’s I test highlights significant spatial autocorrelation in the residuals. This indicates that the model fails to fully account for spatial dependencies.

**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{Response here.}
**Disclosure of assistance:**  
In addition to class materials, I referred to relevant material on Stack Overflow, such as plotting multiple maps with `tmap_arrange` and creating spatial weights matrices using `spdep`. These resources provided practical examples and solutions that helped me better understand the implementation of these tasks. 

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{Discussion here.}
**Lab Reflection:**  
The difficulty of this lab is moderate. Tasks such as handling spatial data, performing Moran's I tests, and interpreting spatial autocorrelation were more challenging and required additional effort to understand. The biggest things I learned from completing this lab include how to analyze spatial patterns in regression residuals and the importance of spatial autocorrelation.

**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}
# Calculate first difference variables (2020 - 2000)
lab6_data <- lab6_data %>%
  mutate(
    diff_ndvi = ndvi_20_med - ndvi_00_med,
    diff_maxtemp = maxtemp_20_med - maxtemp_00_med,
    diff_mintemp = mintemp_20_med - mintemp_00_med,
    diff_rain = rain_20_sum - rain_00_sum,
    diff_pop = pop_20 - pop_00,
    diff_water = water_20_pct - water_00_pct
  )

# Fit a multivariate regression model using the first difference variables
diff_model <- lm(diff_ndvi ~ diff_maxtemp + diff_rain + diff_pop + diff_water, data = lab6_data)

# Summary of the model
summary(diff_model)

# Add residuals to the dataset
lab6_data$diff_residuals <- residuals(diff_model)

# Plot a histogram of residuals
hist(
  lab6_data$diff_residuals,
  main = "Histogram of Residuals (First Difference Model)",
  xlab = "Residuals",
  col = "lightblue",
  border = "white"
)

# Plot the residuals spatially
tm_shape(lab6_data) +
  tm_polygons(
    "diff_residuals",
    style = "jenks",
    palette = "RdBu",
    title = "Residuals (First Difference Model)",
    midpoint = 0
  ) +
  tm_layout(
    main.title = "Spatial Distribution of Residuals",
    legend.outside = TRUE,
    legend.width = 2
  )

# Perform Moran's I test on residuals
coords <- st_coordinates(st_centroid(lab6_data))
nb <- knn2nb(knearneigh(coords, k = 5))
lw <- nb2listw(nb, style = "W")

moran_test_diff <- moran.test(lab6_data$diff_residuals, lw)
print(moran_test_diff)
```
1. The model suggests that changes in NDVI over time are primarily driven by changes in rainfall (positive impact) and maximum temperature (negative impact), while population and water changes appear to have negligible effects. The low adjusted $R^2$ value indicates that other unmeasured factors or processes may also influence NDVI changes, and the model captures only a small portion of the variability. 
2. The histogram shows that most residuals are tightly concentrated around 0, indicating the model's predictions are reasonably close to the observed changes in NDVI. 
3. The Moran's I statistic of 0.2816 indicates significant positive spatial autocorrelation in the residuals of the first difference model. This means that areas with similar residuals are spatially clustered rather than randomly distributed. 

#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

